### 2025-01-23 paper/related.qmd
bibliography sorted and everything adapted to ECML
noice

@@

@teney2020learning propose an approach using counterfactuals in training that does not rely on data augmentation: they argue that counterfactual pairs typically already exist in training datasets. Specifically, their approach relies on, firstly, identifying similar input samples with different annotations and, secondly, ensuring that the gradient of the classifier aligns with the vector between pairs of counterfactual inputs using the cosine distance as a loss function (referred to as *gradient supervision*)

> This might be useful for our task as well...

In the natural language processing (NLP) domain, counterfactuals have similarly been used to improve models through data augmentation: \cite{wu2021polyjuice}, propose POLYJUICE, a general-purpose counterfactual generator for language models. They demonstrate empirically that augmenting training data through POLYJUICE counterfactuals improves robustness in a number of NLP tasks.

@@

@freiesleben2022intriguing propose two definitional differences between Adversarial Examples (AE) and Counterfactual Explanations (CE): firstly, and more importantly according to the authors, the term AE implies missclassification, which is not the case for CE;

> this might be a useful notion for use to distinguish between adversarials and explanations during training

secondly, they argue that closeness plays a more critical role in the context of CE but confess that even counterfactuals that are not close might be relevant explanations. @pawelczyk2022exploring show that CE and AE are equivalent under certain conditions and derive upper bounds on the distances between them.

@@

@guo2023counternet are the first to propose end-to-end training pipeline that includes counterfactual explanations as part of the training prodeduce. In particular, they propose a specific network architecture that includes a predictor and CE generator network,

> akin a GAN?

where the parameters of the CE generator network are learnable. Counterfactuals are generated during each training iteration and fed back to the predictor network (\textcolor{red}{here we are aligned}). In contrast, we impose no restrictions on the neural network architecture at all.

> to ensure the one-hot encoding of categorical features is maintained, they simple use softmax (might be interesting for CE.jl)

Interestingly, the authors find that their approach is sensitive to the choice of the loss function: only MSE seems to lead to good performance. They also demonstrate theoretically, that the objective function is difficult to optimize due to divergent gradients and suffers from poor adversarial robustness.

> because partial gradients with respect to the classification loss component and the counterfactual validity component point in opposite directions

To mitigate these issues, the authors use block-wise gradient descent: they first update with respect to classification loss and then use a second update with respect to the other loss components

> this might be useful for our task as well

@ross2021learning propose a way to train models that are guaranteed to provide recourse for individuals with high probability. The approach builds on adversarial training,

> here we are aligned

where in this context adversarial examples are actively encouraged to exist, but only target attacks with respect to the positive class. The proposed method allows for imposing a set of actionable recourse ex-ante: for example, users can impose mutability constraints for features.

> here we are aligned

> To solve their objective function more efficiently, they use a first-order Taylor approximation to approximate the recourse loss component (might be applicable in our case

>>>>

[@wachter2017counterfactual;@joshi2019realistic;@altmeyer2024faithful]

@@

@teney2020learning propose an approach using counterfactuals in training that does not rely on data augmentation: they argue that counterfactual pairs typically already exist in training datasets. Specifically, their approach relies on, firstly, identifying similar input samples with different annotations and, secondly, ensuring that the gradient of the classifier aligns with the vector between pairs of counterfactual inputs using the cosine distance as a loss function (referred to as *gradient supervision*) (***this might be useful for our task as well***). In the natural language processing (NLP) domain, counterfactuals have similarly been used to improve models through data augmentation: @wu2021polyjuice, propose POLYJUICE, a general-purpose counterfactual generator for language models. They demonstrate empirically that augmenting training data through POLYJUICE counterfactuals improves robustness in a number of NLP tasks.

@@

@freiesleben2022intriguing propose two definitional differences between Adversarial Examples (AE) and Counterfactual Explanations (CE): firstly, and more importantly according to the authors, the term AE implies missclassification, which is not the case for CE (***this might be a useful notion for use to distinguish between adversarials and explanations during training***); secondly, they argue that closeness plays a more critical role in the context of CE but confess that even counterfactuals that are not close might be relevant explanations. @pawelczyk2022exploring show that CE and AE are equivalent under certain conditions and derive upper bounds on the distances between them.

@@

@guo2023counternet are the first to propose end-to-end training pipeline that includes counterfactual explanations as part of the training prodeduce. In particular, they propose a specific network architecture that includes a predictor and CE generator network (***akin a GAN?***), where the parameters of the CE generator network are learnable. Counterfactuals are generated during each training iteration and fed back to the predictor network (***here we are aligned***). In contrast, we impose no restrictions on the neural network architecture at all. (***to ensure the one-hot encoding of categorical features is maintained, they simple use softmax (might be interesting for CE.jl)***) Interestingly, the authors find that their approach is sensitive to the choice of the loss function: only MSE seems to lead to good performance. They also demonstrate theoretically, that the objective function is difficult to optimize due to divergent gradients and suffers from poor adversarial robustness. (***because partial gradients with respect to the classification loss component and the counterfactual validity component point in opposite directions***). To mitigate these issues, the authors use block-wise gradient descent: they first update with respect to classification loss and then use a second update with respect to the other loss components (***this might be useful for our task as well***). @ross2021learning propose a way to train models that are guaranteed to provide recourse for individuals with high probability. The approach builds on adversarial training (***here we are aligned***), where in this context adversarial examples are actively encouraged to exist, but only target attacks with respect to the positive class. The proposed method allows for imposing a set of actionable recourse ex-ante: for example, users can impose mutability constraints for features (***here we are aligned***). (***To solve their objective function more efficiently, they use a first-order Taylor approximation to approximate the recourse loss component (might be applicable in our case)***)

### 2025-01-23 paper/related.tex
noice

\cite{teney2020learning} propose an approach using counterfactuals in
training that does not rely on data augmentation: they argue that
counterfactual pairs typically already exist in training datasets.
Specifically, their approach relies on, firstly, identifying similar
input samples with different annotations and, secondly, ensuring that
the gradient of the classifier aligns with the vector between pairs of
counterfactual inputs using the cosine distance as a loss function
(referred to as \emph{gradient supervision})

\begin{quote}
This might be useful for our task as well\ldots{}
\end{quote}

In the natural language processing (NLP) domain, counterfactuals have
similarly been used to improve models through data augmentation:
\cite{wu2021polyjuice}, propose POLYJUICE, a general-purpose
counterfactual generator for language models. They demonstrate
empirically that augmenting training data through POLYJUICE
counterfactuals improves robustness in a number of NLP tasks.

@@

\cite{freiesleben2022intriguing} propose two definitional differences
between Adversarial Examples (AE) and Counterfactual Explanations (CE):
firstly, and more importantly according to the authors, the term AE
implies missclassification, which is not the case for CE;

\begin{quote}
this might be a useful notion for use to distinguish between
adversarials and explanations during training
\end{quote}

secondly, they argue that closeness plays a more critical role in the
context of CE but confess that even counterfactuals that are not close
might be relevant explanations. \cite{pawelczyk2022exploring} show that
CE and AE are equivalent under certain conditions and derive upper
bounds on the distances between them.

@@

\cite{guo2023counternet} are the first to propose end-to-end training
pipeline that includes counterfactual explanations as part of the
training prodeduce. In particular, they propose a specific network
architecture that includes a predictor and CE generator network,

\begin{quote}
akin a GAN?
\end{quote}

where the parameters of the CE generator network are learnable.
Counterfactuals are generated during each training iteration and fed
back to the predictor network (\textcolor{red}{here we are aligned}). In
contrast, we impose no restrictions on the neural network architecture
at all.

\begin{quote}
to ensure the one-hot encoding of categorical features is maintained,
they simple use softmax (might be interesting for CE.jl)
\end{quote}

Interestingly, the authors find that their approach is sensitive to the
choice of the loss function: only MSE seems to lead to good performance.
They also demonstrate theoretically, that the objective function is
difficult to optimize due to divergent gradients and suffers from poor
adversarial robustness.

\begin{quote}
because partial gradients with respect to the classification loss
component and the counterfactual validity component point in opposite
directions
\end{quote}

To mitigate these issues, the authors use block-wise gradient descent:
they first update with respect to classification loss and then use a
second update with respect to the other loss components

\begin{quote}
this might be useful for our task as well
\end{quote}

\cite{ross2021learning} propose a way to train models that are
guaranteed to provide recourse for individuals with high probability.
The approach builds on adversarial training,

\begin{quote}
here we are aligned
\end{quote}

where in this context adversarial examples are actively encouraged to
exist, but only target attacks with respect to the positive class. The
proposed method allows for imposing a set of actionable recourse
ex-ante: for example, users can impose mutability constraints for
features.

\begin{quote}
here we are aligned
\end{quote}

\begin{quote}
To solve their objective function more efficiently, they use a
first-order Taylor approximation to approximate the recourse loss
component (might be applicable in our case
\end{quote}

>>>>

\cite{teney2020learning} propose an approach using counterfactuals in
training that does not rely on data augmentation: they argue that
counterfactual pairs typically already exist in training datasets.
Specifically, their approach relies on, firstly, identifying similar
input samples with different annotations and, secondly, ensuring that
the gradient of the classifier aligns with the vector between pairs of
counterfactual inputs using the cosine distance as a loss function
(referred to as \emph{gradient supervision}) (\textbf{\emph{this might
be useful for our task as well}}). In the natural language processing
(NLP) domain, counterfactuals have similarly been used to improve models
through data augmentation: \cite{wu2021polyjuice}, propose POLYJUICE, a
general-purpose counterfactual generator for language models. They
demonstrate empirically that augmenting training data through POLYJUICE
counterfactuals improves robustness in a number of NLP tasks.

@@

\cite{freiesleben2022intriguing} propose two definitional differences
between Adversarial Examples (AE) and Counterfactual Explanations (CE):
firstly, and more importantly according to the authors, the term AE
implies missclassification, which is not the case for CE
(\textbf{\emph{this might be a useful notion for use to distinguish
between adversarials and explanations during training}}); secondly, they
argue that closeness plays a more critical role in the context of CE but
confess that even counterfactuals that are not close might be relevant
explanations. \cite{pawelczyk2022exploring} show that CE and AE are
equivalent under certain conditions and derive upper bounds on the
distances between them.

@@

\cite{guo2023counternet} are the first to propose end-to-end training
pipeline that includes counterfactual explanations as part of the
training prodeduce. In particular, they propose a specific network
architecture that includes a predictor and CE generator network
(\textbf{\emph{akin a GAN?}}), where the parameters of the CE generator
network are learnable. Counterfactuals are generated during each
training iteration and fed back to the predictor network
(\textbf{\emph{here we are aligned}}). In contrast, we impose no
restrictions on the neural network architecture at all.
(\textbf{\emph{to ensure the one-hot encoding of categorical features is
maintained, they simple use softmax (might be interesting for CE.jl)}})
Interestingly, the authors find that their approach is sensitive to the
choice of the loss function: only MSE seems to lead to good performance.
They also demonstrate theoretically, that the objective function is
difficult to optimize due to divergent gradients and suffers from poor
adversarial robustness. (\textbf{\emph{because partial gradients with
respect to the classification loss component and the counterfactual
validity component point in opposite directions}}). To mitigate these
issues, the authors use block-wise gradient descent: they first update
with respect to classification loss and then use a second update with
respect to the other loss components (\textbf{\emph{this might be useful
for our task as well}}). \cite{ross2021learning} propose a way to train
models that are guaranteed to provide recourse for individuals with high
probability. The approach builds on adversarial training
(\textbf{\emph{here we are aligned}}), where in this context adversarial
examples are actively encouraged to exist, but only target attacks with
respect to the positive class. The proposed method allows for imposing a
set of actionable recourse ex-ante: for example, users can impose
mutability constraints for features (\textbf{\emph{here we are
aligned}}). (\textbf{\emph{To solve their objective function more
efficiently, they use a first-order Taylor approximation to approximate
the recourse loss component (might be applicable in our case)}})
